Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-18 cordillera-cycle.tex
Prepare for future figure improvements.

During maximum glaciation, the main meridional ice divide is positioned over
the western flank of the Rocky Mountains (Fig.~\ref{fig:snapshots}, lower
panels; Fig.~\ref{fig:mis2}). This result appears to contrast with
palaeoglaciological reconstructions for central and southern British Columbia
that inferred the ice divide to have been situated in a more westerly position,
over the western margin of the Interior Plateau \citep{Ryder.etal.1991,
Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A
latitudinal saddle has been inferred to have connected ice dispersal centres in
the Columbia Mountains with the main ice divide \citep{Ryder.etal.1991,
Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal saddle
does indeed features in our modelling results, however, in an inverse
configuration between the main ice divide over the Columbia Mountains and a
secondary divide over the southern Coast Mountains (Fig.~\ref{fig:mis2}).

@@

Modelled patterns of ice sheet retreat are relatively consistent between the
two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
southern sector of the modelling domain, including the Puget Lowland, the Coast
and Rocky mountains, and the Interior Plateau of British Columbia, becomes
completely deglaciated by 12\,ka, whereas a significant ice cover remains over
the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias mountains in the
northern sector of the modelling domain. After 12\,ka, deglaciation continues
to proceed across the Liard Lowland with a radial ice margin retreat towards
the surrounding mountain ranges, consistent with the regional melt water record
of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
decay by retreating towards the Selwyn and Skeena mountains. The last remnants
of the Cordilleran ice sheet finally disappear from the Skeena mountains around
10.6\,ka (EPICA) and 9.4\,ka (GRIP).

@@

Because a general tenant in glacial geomorphology is that the majority of
landforms (lineations and eskers) are part of the deglacial envelop
\citep[terminology from][]{Kleman.etal.2006}, that is that these were formed
close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
immediately preceding deglaciation or at the time of cessation of sliding
inside the cold-base retreating margin (Fig.~\ref{fig:lastflow}). The
modelled deglacial flow patterns appear consistent between the GRIP and EPICA
simulations. They depict an active ice sheet retreat in the peripheral areas,
followed by stagnant ice decay in some of the interior regions. Several parts
of the modelling domain do not experience any basal sliding throughout the
deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
includes parts of the Interior Plateau in central British Columbia, major
portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
the MacKenzie Mountains.

@@

However, on the Interior Plateau of British Columbia, both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but an impressive set of glacial delineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.
The Interior Plateau lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow patterns over the Interior
Plateau. Another explanation for the incongruent results could be that the
Interior Plateau lineation system predates deglaciation, and that deglacial
landforms are largely absent \footnote{
    \textbf{Martin, Johan}, any southwards running eskers on the plateau?}.

The
modelled deglaciation of the Interior Plateau, central British Columbia,
consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
southwards-flowing (Fig.~\ref{fig:lastflow}) non-sliding ice lobes positioned
in-between deglaciated mountain ranges
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
appears compatible with the prevailing conceptual model of deglaciation of
central British Columbia, in which mountain ranges emerge from the ice before
the plateau \citep[Fig.~7]{Fulton.1991}.

@@

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} (21.4\,cal\,ka)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
           the Skeena Mountains (SM), the Coast Mountains (CM), the Rocky
           Mountains (RM) and the North Cascades (NC). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets and mountain ranges on the map.
                 Mark location of the Puget Lowland.}}
  \label{fig:locmap}
\end{figure}

% fig:atm
\begin{figure}
  \includegraphics{atm}
  \caption{Monthly mean near-surface air temperature, precipitation and
           standard deviation of daily mean temperature for January and July
           months from the North American Regional Reanalysis (NARR)
           climatology, used to force the ice sheet model. Note the
           strong contrasts in seasonality, timing of the precipitation peak,
           and temperature variability over the model domain, notably between
           the maritime and continental regions.
           \todo{Dash boundaries between ice sheets?}}
  \label{fig:atm}
\end{figure}

@@

% fig:duration
\begin{figure*}
  \includegraphics{duration}
  \caption{Modelled duration of ice cover during the last 120\,ka.
           Note the irregular colour scale. A contiguous ice cover spanning
           from the Alaska Range (AR) to the southern Coast Mountains (CM) and
           Rocky Mountains (RM) exists for about 29\,ka in both
           simulations. A central
           ice cover persists over the Skeena Mountains (SM) during most of the
           simulation. On the other hand, the maximal extent of the ice sheet
           generally corresponds to relatively short durations of ice cover.
           \todo{Add WSE, SMK, NR, NC mountain ranges, and changes kyr to ka}}
  \label{fig:duration}
\end{figure*}

@@

% fig:warmfrac
\begin{figure*}
  \includegraphics{warmfrac}
  \caption{Modelled fraction of warm-based ice cover during the ice-covered
           period. Note the dominance of warm-based conditions on the
           continental shelf and major glacial troughs of the coastal ranges.
           Hatches indicate areas that were covered by cold ice only.
           \todo{indicate location of the Skeena Mountains.}}
  \label{fig:warmfrac}
\end{figure*}

@@

% fig:deglacshots
\begin{figure*}
  \includegraphics{deglacshots}
  \caption{Snapshots of modelled surface topography (200\,m contours)
           and surface velocity (colour mapping) during the last deglaciation
           from the GRIP (top panels) and EPICA (bottom panels) simulations.}
  \label{fig:deglacshots}
\end{figure*}

@@

% fig:lastflow
\begin{figure*}
  \includegraphics{lastflow}
  \caption{Modelled directions of the deglacial basal flow velocities. Hatches
           indicate areas that remain non-sliding throughout deglaciation
           (22.0--8.0\,ka), notably including the Interior Plateau of central
           British Columbia.
           Sliding grid cells were distinguished from non-sliding grid cells
           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.}
  \label{fig:lastflow}
\end{figure*}

% fig:profiles-grip
\begin{figure}
  \includegraphics{profiles-grip}
  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
           during deglaciation (22.0--8.0\,ka) in the GRIP 6\,km
           simulation, corresponding to the four transects indicated in
           Fig.~\ref{fig:deglac}.}
  \label{fig:profiles-grip}
\end{figure}

% fig:profiles-epica
\begin{figure}
  \includegraphics{profiles-epica}
  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
           during deglaciation (22.0--8.0\,ka) in the EPICA 6\,km
           simulation, corresponding to the four transects indicated in
           Fig.~\ref{fig:deglac}.}
  \label{fig:profiles-epica}
\end{figure}

>>>>

During maximum glaciation, the main meridional ice divide is positioned over
the western flank of the Rocky Mountains (Fig.~\ref{fig:snapshots}, lower
panels; Fig.~\ref{fig:mis2}). This result appears to contrast with
palaeoglaciological reconstructions for central and southern British Columbia
that inferred the ice divide to have been situated in a more westerly position,
over the western margin of the Fraser Plateau\footnote{
    Just checked that this name is correct. Feels lighter than the ``Interior
    Plateau of south-central British Columbia'' (although having a slightly
    different meaning).}
\citep{Ryder.etal.1991,
Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A
latitudinal saddle has been inferred to have connected ice dispersal centres in
the Columbia Mountains with the main ice divide \citep{Ryder.etal.1991,
Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal saddle
does indeed features in our modelling results, however, in an inverse
configuration between the main ice divide over the Columbia Mountains and a
secondary divide over the southern Coast Mountains (Fig.~\ref{fig:mis2}).

@@

Modelled patterns of ice sheet retreat are relatively consistent between the
two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
southern sector of the modelling domain, including the Puget Lowland, the Coast
and Rocky mountains, and the Fraser Plateau, becomes
completely deglaciated by 12\,ka, whereas a significant ice cover remains over
the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias mountains in the
northern sector of the modelling domain. After 12\,ka, deglaciation continues
to proceed across the Liard Lowland with a radial ice margin retreat towards
the surrounding mountain ranges, consistent with the regional melt water record
of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
decay by retreating towards the Selwyn and Skeena mountains. The last remnants
of the Cordilleran ice sheet finally disappear from the Skeena mountains around
10.6\,ka (EPICA) and 9.4\,ka (GRIP).

@@

Because a general tenant in glacial geomorphology is that the majority of
landforms (lineations and eskers) are part of the deglacial envelop
\citep[terminology from][]{Kleman.etal.2006}, that is that these were formed
close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
immediately preceding deglaciation or at the time of cessation of sliding
inside the cold-base retreating margin (Fig.~\ref{fig:lastflow}). The
modelled deglacial flow patterns appear consistent between the GRIP and EPICA
simulations. They depict an active ice sheet retreat in the peripheral areas,
followed by stagnant ice decay in some of the interior regions. Several parts
of the modelling domain do not experience any basal sliding throughout the
deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
includes parts of the Fraser Plateau, major
portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
the MacKenzie Mountains.

@@

However, on the Fraser Plateau, both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but an impressive set of glacial delineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow patterns over the Interior
Plateau. Another explanation for the incongruent results could be that the
Fraser Plateau lineation system predates deglaciation, and that deglacial
landforms are largely absent \footnote{
    \textbf{Martin, Johan}, any southwards running eskers on the plateau?}.

The modelled deglaciation of the Fraser Plateau
consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
southwards-flowing (Fig.~\ref{fig:lastflow}) non-sliding ice lobes positioned
in-between deglaciated mountain ranges
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
appears compatible with the prevailing conceptual model of deglaciation of
central British Columbia, in which mountain ranges emerge from the ice before
the plateau \citep[Fig.~7]{Fulton.1991}.

@@

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} (21.4\,cal\,ka)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
           the Skeena Mountains (SM), the Coast Mountains (CM), the Rocky
           Mountains (RM) and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Fraser Plateau (FP) and the
           Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets, mountain ranges and depressions on
                 the map. Dash boundaries between ice sheets?}}
  \label{fig:locmap}
\end{figure}

% fig:atm
\begin{figure}
  \includegraphics{atm}
  \caption{Monthly mean near-surface air temperature, precipitation and
           standard deviation of daily mean temperature for January and July
           months from the North American Regional Reanalysis (NARR)
           climatology, used to force the ice sheet model. Note the
           strong contrasts in seasonality, timing of the precipitation peak,
           and temperature variability over the model domain, notably between
           the maritime and continental regions.}
  \label{fig:atm}
\end{figure}

@@

% fig:duration
\begin{figure*}
  \includegraphics{duration}
  \caption{Modelled duration of ice cover during the last 120\,ka.
           Note the irregular colour scale. A contiguous ice cover spanning
           from the Alaska Range (AR) to the southern Coast Mountains (CM) and
           Rocky Mountains (RM) exists for about 29\,ka in both
           simulations. A central
           ice cover persists over the Skeena Mountains (SM) during most of the
           simulation. On the other hand, the maximal extent of the ice sheet
           generally corresponds to relatively short durations of ice cover.
           \todo{Add WSE, SMK, NR, NC mountain ranges, and change kyr to ka}}
  \label{fig:duration}
\end{figure*}

@@

% fig:warmfrac
\begin{figure*}
  \includegraphics{warmfrac}
  \caption{Modelled fraction of warm-based ice cover during the ice-covered
           period. Note the dominance of warm-based conditions on the
           continental shelf and major glacial troughs of the coastal ranges.
           Hatches indicate areas that were covered by cold ice only.
           \todo{indicate location of the Skeena Mountains and the MacKenzie
                 Mountains.}}
  \label{fig:warmfrac}
\end{figure*}

@@

% fig:deglacshots
\begin{figure*}
  \includegraphics{deglacshots}
  \caption{Snapshots of modelled surface topography (200\,m contours)
           and surface velocity (colour mapping) during the last deglaciation
           from the GRIP (top panels) and EPICA (bottom panels) simulations.
           \todo{Add the location of profiles used in
                 Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-grip}.}}
  \label{fig:deglacshots}
\end{figure*}

@@

% fig:lastflow
\begin{figure*}
  \includegraphics{lastflow}
  \caption{Modelled directions of the deglacial basal flow velocities. Hatches
           indicate areas that remain non-sliding throughout deglaciation
           (22.0--8.0\,ka), notably including the Fraser Plateau.
           Sliding grid cells were distinguished from non-sliding grid cells
           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.
           \todo{indicate the Fraser Plateau and the Liard Lowland. Add the
                 location of profiles used in Figs.~\ref{fig:profiles-grip}
                 and~\ref{fig:profiles-grip}.}}
  \label{fig:lastflow}
\end{figure*}

% fig:profiles-grip
\begin{figure}
  \includegraphics{profiles-grip}
  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
           during deglaciation (22.0--8.0\,ka) in the GRIP 6\,km
           simulation, corresponding to the four transects indicated in
           Fig.~\ref{fig:deglac}.
           \todo{indicate mountain ranges and depressions.}}
  \label{fig:profiles-grip}
\end{figure}

% fig:profiles-epica
\begin{figure}
  \includegraphics{profiles-epica}
  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
           during deglaciation (22.0--8.0\,ka) in the EPICA 6\,km
           simulation, corresponding to the four transects indicated in
           Fig.~\ref{fig:deglac}.
           \todo{indicate mountain ranges and depressions.}}
  \label{fig:profiles-epica}
\end{figure}
